---
title: "Cycle model: Get molecular model day"
author: "Jessica Chung"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: true
    toc_depth: 4
---

This notebook gets an estimated day for all samples with expression data.

It first performs batch correction using ComBat to correct for batch effects.

Next, it re-assigns proliferative samples into early, mid, and late proliferative by fitting a spline model using menstrual, proliferative, and early secretory samples. This uses the assumption that proliferative samples were approximately drawn from a uniform distribution, so the number of early, mid, and late proliferative samples would be approximately equal.

Then, using the 7 stages, a cyclic cublic spline is fit.  

Then it assigns a model 'day' using the time which minimises the mean squared error between the observed expression data and the gene models. This 'day' is just a relative timepoint in the cycle, and should not be seen as corresponding to a real day.

Then it transforms the days so they're uniformly distributed within each stage, using the number of samples in the pathology dating to anchor the relative days back to something that resembles the original pathology groups. This uses the assumption that the original sample distribution from each stage was mostly uniform.

Keep in mind: days in the molecular model may not map linearly to uterine days in reality, so drawing conclusions based on molecular days may be a bit dubious.


```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=TRUE, message=TRUE, error=TRUE, echo=TRUE, results="hold")
knitr::opts_knit$set(root.dir = "..")
options(digits=4)
options(width=120)
Sys.setenv(TZ = "Australia/Melbourne")
```

-----

# Load

```{r message=FALSE}
library(tidyverse)
library(edgeR)
library(limma)
library(ggplot2)
library(mgcv)
library(splines)
```

```{r}
qld_counts_list <- readRDS("data/tidy_data/qld_counts_list.rds")
sample_info <- readRDS("data/tidy_data/sample_info.rds")
all_annotations <- readRDS("data/tidy_data/all_annotations.rds")
combat_nc <- readRDS("data/tidy_data/batch_normalised_exprs.rds")
```

```{r}
sample_info <- sample_info %>% 
  mutate(study=ifelse(str_detect(sample_id, "^URS"), "study_2", "study_1")) %>%
  filter(! is.na(rna_seq_id))
```

-----

# Redefine stages

```{r}
# Which samples to use when fitting the models?
table(sample_info$cycle_stage, useNA="always")

# Only use study 1 samples
pheno <- sample_info %>% 
  filter(cycle_stage %in% 1:7,
         sample_id %in% colnames(combat_nc),
         study == "study_1") %>%
  select(sample_id, cycle_stage, study)
pheno$cycle_stage %>% table
```

-----

# Spline model

```{r}
fit_spline_model <- function(y, x, spline_k=6, spline_fx=FALSE, spline_bs="cc", 
                             gamma=1, predict_range=seq(0,28,by=0.1), 
                             weights=NULL, knots=NULL, return_coefs=FALSE,
                             return_fit=TRUE) {
  
  fit <- gam(y ~ s(x, bs=spline_bs, k=spline_k, fx=spline_fx), 
             knots=knots, weights=weights, gamma=gamma)
  fit_summary <- summary(fit)
  
  pred <- predict(fit, newdata=data.frame(x=predict_range))
  pred <- data.frame(cycle_time=predict_range, pred=pred)
  
  results <- list(pred=pred, 
                  edf=round(fit_summary$edf, 3),
                  R2=round(fit_summary$r.sq, 4),
                  dev_exp=round(fit_summary$dev.expl, 4),
                  s_table=fit_summary$s.table)
  if (return_coefs) results$coefs <- fit$coefficients
  if (return_fit) results$fit <- fit
  return(results)
}

plot_spline_model <- function(spline_list, y, x, gene_name="", alpha_point=1,
                              alpha_line=0.5, jitter_height=0, jitter_width=0.05, 
                              plot=TRUE) {
  dat <- data.frame(exprs=y, cycle_time=x)
  g <- ggplot(dat, aes(x=cycle_time, y=exprs)) +
    geom_jitter(height=jitter_height, width=jitter_width, alpha=alpha_point) +
    geom_line(spline_list$pred, mapping=aes(x=cycle_time, y=pred), alpha=alpha_line) +
    labs(title=sprintf("%s | R^2 = %0.3f | edf = %0.2f", 
                       gene_name, spline_list$R2, spline_list$edf))
  if (! plot) return(g)
  print(g)
}
```

-----

# Assign proliferative stages

According to Peter, it's safe to assume that the distribution of samples in the 
proliferative stage is about uniform.

Use the first round of fitting the spline model to separate proliferative stage 
samples into equal groups of early, mid, and late proliferative.

```{r}
# Set all proliferative samples as day 10 (stage 3)
pheno <- pheno %>%
  mutate(cycle_stage_reassigned_prolif=case_when(
    cycle_stage == "2" ~ "3",
    cycle_stage == "4" ~ "3",
    TRUE ~ cycle_stage
  ))
table(pheno$cycle_stage_reassigned_prolif)
```

## Fit spline

```{r}
subset_pheno <- pheno %>% filter(cycle_stage_reassigned_prolif %in% c(1,3,5))
subset_cycle_time <- as.numeric(subset_pheno$cycle_stage_reassigned_prolif)
subset_cycle_range <- seq(1, 5, 0.1)

exprs <- combat_nc[,subset_pheno$sample_id]
genes <- rownames(combat_nc)
gene_models <- lapply(genes, function(p) {
  fit_spline_model(y=exprs[p,], x=subset_cycle_time, 
                   spline_k=3, gamma=4, spline_bs="cr",
                   predict_range=subset_cycle_range, weights=NULL,
                   return_fit=FALSE)
})
names(gene_models) <- genes
```


```{r}
# Get expected values for each cycle day for each probe
expected_exprs <- sapply(genes, function(p) gene_models[[p]]$pred$pred) %>% 
  t %>% magrittr::set_colnames(paste0("time_", subset_cycle_range))

# Get probe R^2 values
r2 <- sapply(genes, function(p) gene_models[[p]]$R2)

# Get df for models
spline_edf <- sapply(genes, function(p) gene_models[[p]]$edf)
table(cut(spline_edf, breaks = 5))

# Get deviance explained
dev_exp <- sapply(genes, function(p) gene_models[[p]]$dev_exp)
cor(r2, dev_exp)

# Get coefficients
#coefs <- sapply(genes, function(p) gene_models[[p]]$coefs) %>% t
#colnames(coefs) <- str_remove_all(colnames(coefs), "[()]")

# Get p-values
pvals <- sapply(genes, function(p) gene_models[[p]]$s_table[1,"p-value"])

model_df <- data.frame(probe_id=genes, r2=r2, spline_edf=spline_edf, 
                       dev_exp=dev_exp, p=pvals,
                       stringsAsFactors=FALSE)
```


```{r}
ggplot(model_df %>% head(1000), aes(x=spline_edf, y=r2)) +
  geom_point(alpha=0.2, size=1)

ggplot(model_df %>% head(1000), aes(x=p, y=r2)) +
  geom_point(alpha=0.2, size=1) +
  scale_x_log10()
```

r2, deviance explained, and p-values are highly correlated

```{r results="markup"}
# Examine some genes

# Linear
g <- model_df %>% filter(spline_edf %>% between(1, 1)) %>% head(1)
g %>% t
plot_spline_model(gene_models[[g$probe_id]], x=subset_cycle_time, y=exprs[g$probe_id,], gene_name=g$probe_id)

# Medium spline_edf but low r2
g <- model_df %>% filter(spline_edf > 1.5, r2 < 0.1) %>% head(1)
g %>% t
plot_spline_model(gene_models[[g$probe_id]], x=subset_cycle_time, y=exprs[g$probe_id,], gene_name=g$probe_id)

# Okay fit
g <- model_df %>% filter(r2 %>% between(0.20, 0.22)) %>% head(1)
g %>% t
plot_spline_model(gene_models[[g$probe_id]], x=subset_cycle_time, y=exprs[g$probe_id,], gene_name=g$probe_id)

# High r2
g <- model_df %>% arrange(desc(r2)) %>% head(1)
g %>% t
plot_spline_model(gene_models[[g$probe_id]], x=subset_cycle_time, y=exprs[g$probe_id,], gene_name=g$probe_id)
```

## Get best time estimate

```{r}
# Get best estimate
fit1_best_estimate <- sapply(subset_pheno$sample_id, function(s) {
  sweep(x=expected_exprs, MARGIN=1, STATS=combat_nc[,s]) %>% 
    `^`(2) %>% colSums %>% which.min %>% names %>% str_replace("time_", "") %>%
    as.numeric
})
```


```{r}
subset_pheno <- subset_pheno %>%
  mutate(best_estimate=fit1_best_estimate[sample_id])

ggplot(subset_pheno, aes(x=cycle_stage, y=best_estimate, color=study)) +
  geom_jitter(height=0, width=0.2) +
  labs(x="Cycle stage from pathology", 
       y="Best estimate of cycle time using model",
       title="Pathology cycle stage vs molecular model cycle time prediction")


ggplot(subset_pheno, aes(x=cycle_stage_reassigned_prolif, y=best_estimate, color=study)) +
  geom_jitter(height=0, width=0.2) +
  theme_bw() +
  labs(x="Cycle stage from pathology", 
       y="Best estimate of cycle time using model",
       title="Pathology cycle stage vs molecular model cycle time prediction")
```

```{r}
saveRDS(subset_pheno, file="cache/fig_2b_data.rds")
```


```{r}
prolif_pheno <- subset_pheno %>% filter(cycle_stage_reassigned_prolif == "3") %>% filter(study == "study_1")

set.seed(42)
prolif_cuts <- cut(rank(prolif_pheno$best_estimate, ties.method="random"), breaks=3)
levels(prolif_cuts) <- c("early", "mid", "late")
table(prolif_cuts)
prolif_pheno$prolif_stage <- prolif_cuts
```

```{r}
saveRDS(prolif_pheno, file="cache/fig_2bii_data.rds")
```

```{r}
fit1_pheno <- pheno
fit1_model_df <- model_df
fit1_gene_models <- gene_models
```

-----

# Fit 7 stages

```{r}
new_cycle_stage <- pheno$cycle_stage
names(new_cycle_stage) <- pheno$sample_id
new_cycle_stage[prolif_pheno %>% filter(prolif_stage == "early") %>% pull(sample_id)] <- 2
new_cycle_stage[prolif_pheno %>% filter(prolif_stage == "mid") %>% pull(sample_id)] <- 3
new_cycle_stage[prolif_pheno %>% filter(prolif_stage == "late") %>% pull(sample_id)] <- 4

pheno <- pheno %>% mutate(new_cycle_stage = new_cycle_stage)
pheno$new_cycle_stage %>% table
```

## Fit spline

```{r}
exprs <- combat_nc[,pheno$sample_id]

# Offset cycle_time by -1 so cyclic spline knots can be placed correctly.
# i.e stage 1 -> 0, stage 7 -> 6
# Don't forget to re-add 1 downstream later!

cycle_time <- as.numeric(pheno$new_cycle_stage) - 1
table(cycle_time)

# Manually set knots
spline_knots <- list(x=0:7)
spline_knots

genes <- rownames(combat_nc)
cycle_range <- seq(0, 7, 0.01)

gene_models <- lapply(genes, function(p) {
  fit_spline_model(y=exprs[p,], x=cycle_time, spline_k=8, gamma=4,
                   predict_range=cycle_range, weights=NULL, knots=spline_knots,
                   return_fit=FALSE)
})
names(gene_models) <- genes
```


```{r}
# Get expected values for each cycle day for each probe
expected_exprs <- sapply(genes, function(p) gene_models[[p]]$pred$pred) %>% 
  t %>% magrittr::set_colnames(paste0("time_", cycle_range))

# Get probe R^2 values
r2 <- sapply(genes, function(p) gene_models[[p]]$R2)

# Get df for models
spline_edf <- sapply(genes, function(p) gene_models[[p]]$edf)
table(cut(spline_edf, breaks = 5))

# Get deviance explained
dev_exp <- sapply(genes, function(p) gene_models[[p]]$dev_exp)
cor(r2, dev_exp)

# Get coefficients
# coefs <- sapply(genes, function(p) gene_models[[p]]$coefs) %>% t
# colnames(coefs) <- str_remove_all(colnames(coefs), "[()]")

# Get p-values
pvals <- sapply(genes, function(p) gene_models[[p]]$s_table[1,"p-value"])

model_df <- data.frame(probe_id=genes, r2=r2, spline_edf=spline_edf, 
                       dev_exp=dev_exp, p=pvals, 
                       stringsAsFactors=FALSE)
```


```{r}
ggplot(model_df, aes(x=spline_edf, y=r2)) +
  geom_point(alpha=0.2, size=1)
```

```{r}
ggplot(model_df, aes(x=dev_exp)) +
  geom_histogram(binwidth=0.02)
```

```{r results="markup"}
# Examine some genes
# Note: these days are offset by 2.5

# Flat
g <- model_df %>% filter(spline_edf %>% between(0, 0)) %>% head(1)
g %>% t
plot_spline_model(gene_models[[g$probe_id]], x=cycle_time, y=exprs[g$probe_id,], gene_name=g$probe_id)

# Almost flat
g <- model_df %>% filter(spline_edf %>% between(0.9, 1.1)) %>% head(1)
g %>% t
plot_spline_model(gene_models[[g$probe_id]], x=cycle_time, y=exprs[g$probe_id,], gene_name=g$probe_id)

# High spline_edf but low r2
g <- model_df %>% filter(spline_edf > 3, r2 < 0.1) %>% head(1)
g %>% t
plot_spline_model(gene_models[[g$probe_id]], x=cycle_time, y=exprs[g$probe_id,], gene_name=g$probe_id)

# Okay fit
g <- model_df %>% filter(spline_edf %>% between(3, 3.5), r2 %>% between(0.25, 0.3)) %>% head(1)
g %>% t
plot_spline_model(gene_models[[g$probe_id]], x=cycle_time, y=exprs[g$probe_id,], gene_name=g$probe_id)

# Good fits: high spline_edf and high r2
g <- model_df %>% filter(spline_edf > 3) %>% arrange(desc(r2)) %>% head(1)
g %>% t
plot_spline_model(gene_models[[g$probe_id]], x=cycle_time, y=exprs[g$probe_id,], gene_name=g$probe_id)
```

## Get best time estimate

```{r}
# Get best estimate
fit2_best_estimate <- sapply(colnames(combat_nc), function(s) {
  sweep(x=expected_exprs, MARGIN=1, STATS=combat_nc[,s]) %>% 
    `^`(2) %>% colSums %>% which.min %>% names %>% str_replace("time_", "") %>%
    as.numeric
})

# Re-add 0.5 so that menstrual is 0 to 1
fit2_best_estimate <- round((fit2_best_estimate + 0.5) %% 7, 2)
```

```{r}
# Generate MSE for future plotting
mse <- sapply(colnames(combat_nc), function(s) {
  sweep(x=expected_exprs, MARGIN=1, STATS=combat_nc[,s]) %>% 
    `^`(2) %>% colMeans
})
```

```{r}
pheno <- pheno %>%
  mutate(best_estimate=fit2_best_estimate[sample_id])

ggplot(pheno, aes(x=cycle_stage_reassigned_prolif , y=best_estimate, color=study)) +
  geom_jitter(height=0, width=0.2) +
  labs(x="Cycle stage from pathology", 
       y="Best estimate of cycle time using model",
       title="Pathology cycle stage vs molecular model cycle day prediction")
```

```{r}
ggplot(pheno, aes(x=new_cycle_stage, y=best_estimate, color=study)) +
  geom_jitter(height=0, width=0.2) +
  labs(x="Cycle stage (re-assigned proliferative)", 
       y="Best estimate of cycle day using model")
```

```{r}
saveRDS(pheno, file="cache/fig2c.rds")
```


```{r}
ggplot(pheno, aes(x=best_estimate, fill=study)) +
  geom_histogram(binwidth=0.1)
```

```{r}
fit2_pheno <- pheno
fit2_model_df <- model_df
fit2_gene_models <- gene_models
```

-----

# Transform time

Transform time to a scale of 0 to 100. 0 should be the start of menstruation.

```{r}
ggplot(fit2_pheno, aes(x=best_estimate, y=cycle_stage_reassigned_prolif)) +
  geom_jitter(height=0.1, width=0) +
  scale_y_discrete(limits = factor(c(7:1)))
```

Cycle stages overlap by quite a bit... 

```{r}
ggplot(fit2_pheno %>% filter(cycle_stage_reassigned_prolif %in% c(1,7)), 
       aes(x=best_estimate, y=cycle_stage_reassigned_prolif)) +
  geom_jitter(height=0.1, width=0) +
  scale_y_discrete(limits = factor(c(7,1))) +
  scale_x_continuous(limits=c(0,2), breaks=0:2)
```

```{r}
fit2_pheno %>% arrange(best_estimate) %>% head(10)
```

First 6 samples are intermixed stage 7, 1, 7, 1, 7, 1. These samples are between 0 and 0.29.

Take the midpoint (0.15) and offset everything by that amount, so the beginning 
is the in the middle of the region with mixed stage 7/1 samples.

Then remap time so the scale is from 0 - 100.

```{r}
offset <- 0.15
pheno <- fit2_pheno %>%
  mutate(best_estimate_offset=(best_estimate - offset) %% 7) %>%
  arrange(best_estimate_offset) %>%
  mutate(rank=1:nrow(fit2_pheno)) %>%
  mutate(transformed_time=(rank - 1)/nrow(fit2_pheno) * 100)
```

```{r}
ggplot(pheno, aes(x=transformed_time, y=cycle_stage_reassigned_prolif)) +
  geom_jitter(height=0.1, width=0) +
  scale_y_discrete(limits = factor(c(7:1))) +
  scale_x_continuous(breaks=seq(0, 100, by=10))
```

```
Approximately:
[0 - 7) : menstrual
[7 - 58) : proliferative
[58 - 100) : secretory
```

```{r}
# Offset categorical stages by 0.5 so they're centred when plotting
dat <- pheno %>%
  mutate(cycle_stage_reassigned_prolif=as.numeric(cycle_stage_reassigned_prolif) - 0.5,
         new_cycle_stage=as.numeric(new_cycle_stage) - 0.5,
         transformed_time=transformed_time/100*7)
ggplot(dat) +
  geom_point(aes(x=cycle_stage_reassigned_prolif, y="initial_cycle_stage")) +
  geom_segment(aes(x=cycle_stage_reassigned_prolif, xend=new_cycle_stage, 
                   y="initial_cycle_stage", yend="reassigned_cycle_time"), size=0.1) +
  geom_point(aes(x=new_cycle_stage, y="reassigned_cycle_time")) +
  geom_segment(aes(x=new_cycle_stage, xend=best_estimate_offset, 
                   y="reassigned_cycle_time", yend="best_estimate_time"), size=0.1) +
  geom_segment(aes(x=best_estimate_offset, xend=transformed_time, 
                   y="best_estimate_time", yend="transformed_time"), size=0.1) +
  scale_x_continuous(breaks=seq(0,7,1)) +
  scale_y_discrete(limits = c("transformed_time", "best_estimate_time", 
                              "reassigned_cycle_time", "initial_cycle_stage")) +
  labs(y="")
```

```{r}
saveRDS(dat, file="cache/fig2d.rds")
```


```{r}
# Create mapping function
f <- splinefun(x=pheno$best_estimate_offset, y=pheno$transformed_time)
```

-----

# Combine phenotype data

```{r}
combat_phenotype <- merge(
  sample_info,
  pheno %>% select(sample_id, cycle_stage_reassigned_prolif:transformed_time),
  by="sample_id", all.x=TRUE)

# Reorder to match combat exprs matrix
m <- match(colnames(combat_nc), combat_phenotype$sample_id)
combat_phenotype <- combat_phenotype[m,]

combat_phenotype <- combat_phenotype %>%
  mutate(model_estimate=fit2_best_estimate[sample_id],
         model_estimate_offset=(model_estimate - offset) %% 7,
         transformed_time_2=f(model_estimate_offset))
```

-----

# Save data

```{r}
unique_id <- "2021-03-23"
saveRDS(combat_phenotype, file=sprintf("data/tidy_data/combat_phenotype_%s.rds", unique_id))

save(fit1_pheno, fit1_model_df, fit2_pheno, fit2_model_df, mse,
     file="cache/1.05_data.RData") 
```

-----

# Session Info

```{r}
devtools::session_info()
```

