---
title: "Cycle model: Normalisation"
author: "Jessica Chung"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: true
    toc_depth: 4
---

This notebook normalises the expression data to remove cycle time effects.

Using the molecular time estimate from the previous notebook, a spline is fit for each
gene, and the curve is subtracted from the expression to obtain a normalised
value.

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=TRUE, message=TRUE, error=TRUE, echo=TRUE, results="hold")
knitr::opts_knit$set(root.dir = "..")
options(digits=4)
options(width=120)
Sys.setenv(TZ = "Australia/Melbourne")
```

-----

# Load

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(sva)
library(splines)
```

```{r}
fit_spline_model <- function(y, x, spline_k=6, spline_fx=FALSE, spline_bs="cc", 
                             gamma=1, predict_range=seq(0,28,by=0.1), 
                             weights=NULL, knots=NULL, return_coefs=FALSE,
                             return_fit=TRUE) {
  
  fit <- gam(y ~ s(x, bs=spline_bs, k=spline_k, fx=spline_fx), 
             knots=knots, weights=weights, gamma=gamma)
  fit_summary <- summary(fit)
  
  pred <- predict(fit, newdata=data.frame(x=predict_range))
  pred <- data.frame(cycle_time=predict_range, pred=pred)
  
  results <- list(pred=pred, 
                  edf=round(fit_summary$edf, 3),
                  R2=round(fit_summary$r.sq, 4),
                  dev_exp=round(fit_summary$dev.expl, 4),
                  s_table=fit_summary$s.table)
  if (return_coefs) results$coefs <- fit$coefficients
  if (return_fit) results$fit <- fit
  return(results)
}


plot_spline_model <- function(spline_list, y, x, gene_name="", alpha_point=1,
                              alpha_line=0.5, jitter_height=0, jitter_width=0.05, 
                              plot=TRUE) {
  dat <- data.frame(exprs=y, cycle_time=x)
  g <- ggplot(dat, aes(x=cycle_time, y=exprs)) +
    geom_jitter(height=jitter_height, width=jitter_width, alpha=alpha_point) +
    geom_line(spline_list$pred, mapping=aes(x=cycle_time, y=pred), alpha=alpha_line) +
    labs(title=sprintf("%s | R^2 = %0.3f | edf = %0.2f", 
                       gene_name, spline_list$R2, spline_list$edf))
  if (! plot) return(g)
  print(g)
}
```

```{r}
# qld_counts_list <- readRDS("data/tidy_data/qld_counts_list.rds")
# sample_info <- readRDS("data/tidy_data/sample_info.rds")
# all_annotations <- readRDS("data/tidy_data/all_annotations.rds")
combat_nc <- readRDS("data/tidy_data/batch_normalised_exprs.rds")
```

```{r}
dataset_id <- "2021-03-23"
combat_phenotype <- readRDS(file=sprintf("data/tidy_data/combat_phenotype_%s.rds", dataset_id))
```

-----

# Correct for cycle stage

Using the transformed day obtained from the previous model.

## Fit spline

```{r}
# Only study 1 samples
pheno <- combat_phenotype %>% filter(! is.na(transformed_time))
exprs <- combat_nc[,pheno$sample_id]
cycle_time <- as.numeric(pheno$transformed_time)
genes <- rownames(exprs)
cycle_range <- seq(0, 100, by=1)

# Manually set knots
k_knots <- 30
spline_knots <- list(x=seq(0,100,length=k_knots))

gene_models <- lapply(genes, function(p) {
  fit_spline_model(y=exprs[p,], x=cycle_time, spline_k=k_knots, gamma=4, knots=spline_knots,
                   predict_range=cycle_range, return_fit=FALSE)
})
names(gene_models) <- genes
```

```{r}
# Get expected values for each cycle day for each probe
expected_exprs <- sapply(genes, function(p) gene_models[[p]]$pred$pred) %>% 
  t %>% magrittr::set_colnames(paste0("time_", cycle_range))

# Get probe R^2 values
r2 <- sapply(genes, function(p) gene_models[[p]]$R2)

# Get df for models
spline_edf <- sapply(genes, function(p) gene_models[[p]]$edf)
table(cut(spline_edf, breaks = 5))

# Get deviance explained
dev_exp <- sapply(genes, function(p) gene_models[[p]]$dev_exp)
cor(r2, dev_exp)

# Get coefficients
# coefs <- sapply(genes, function(p) gene_models[[p]]$coefs) %>% t
# colnames(coefs) <- str_remove_all(colnames(coefs), "[()]")

# Get p-values
pvals <- sapply(genes, function(p) gene_models[[p]]$s_table[1,"p-value"])

model_df <- data.frame(probe_id=genes, r2=r2, spline_edf=spline_edf, 
                       dev_exp=dev_exp, p=pvals, 
                       stringsAsFactors=FALSE)
```

```{r}
ggplot(model_df, aes(x=spline_edf, y=r2)) +
  geom_point(alpha=0.2, size=1)
```

```{r results="markup"}
# Examine some genes

# Flat
g <- model_df %>% filter(spline_edf %>% between(0, 0)) %>% head(1)
g %>% t
plot_spline_model(gene_models[[g$probe_id]], x=cycle_time, y=exprs[g$probe_id,], gene_name=g$probe_id)

# Almost flat
g <- model_df %>% filter(spline_edf %>% between(1, 1.5)) %>% head(1)
g %>% t
plot_spline_model(gene_models[[g$probe_id]], x=cycle_time, y=exprs[g$probe_id,], gene_name=g$probe_id)

# High spline_edf but low r2
g <- model_df %>% filter(spline_edf > 3, r2 < 0.15) %>% head(1)
g %>% t
plot_spline_model(gene_models[[g$probe_id]], x=cycle_time, y=exprs[g$probe_id,], gene_name=g$probe_id)

# Okay fit
g <- model_df %>% filter(spline_edf %>% between(3, 6.5), r2 %>% between(0.30, 0.31)) %>% head(1)
g %>% t
plot_spline_model(gene_models[[g$probe_id]], x=cycle_time, y=exprs[g$probe_id,], gene_name=g$probe_id)

# Good fits: high spline_edf and high r2
g <- model_df %>% arrange(desc(r2)) %>% head(1)
g %>% t
plot_spline_model(gene_models[[g$probe_id]], x=cycle_time, y=exprs[g$probe_id,], gene_name=g$probe_id)
```

```{r}
for (g in model_df %>% arrange(desc(r2)) %>% head(5) %>% pull(probe_id)) {
  plot_spline_model(gene_models[[g]], x=cycle_time, y=exprs[g,], gene_name=g)
}
```

```{r}
# Get best estimate (just to check, don't intend to use)
fit3_best_estimate <- sapply(colnames(combat_nc), function(s) {
  sweep(x=expected_exprs, MARGIN=1, STATS=combat_nc[,s]) %>% 
    `^`(2) %>% colSums %>% which.min %>% names %>% str_replace("time_", "") %>%
    as.numeric
})
```

```{r}
ggplot(data.frame(be1=combat_phenotype$transformed_time_2, be2=fit3_best_estimate),
       aes(x=be1, y=be2)) +
  geom_point(alpha=0.3) +
  geom_abline(color="red", linetype="dashed")
```

-----

# Cycle day correction

```{r}
stopifnot(colnames(combat_nc) == combat_phenotype$sample_id)
cycle_time <- round(combat_phenotype$transformed_time_2)
obs <- combat_nc
exp <- expected_exprs[,paste0("time_", cycle_time)]
gene_means <- apply(exp, 1, mean)
dev <- obs - exp
normalised_exprs <- sweep(dev, MARGIN=1, STATS=gene_means, FUN="+")
```


```{r}
g <- "ENSG00000187231"
plot_spline_model(gene_models[[g]],
                  y=combat_nc[g,], x=cycle_time,
                  gene_name = g, plot=FALSE) +
  scale_y_continuous(limits=c(5.0, 9.1))
dat <- data.frame(cycle_time=cycle_time,
                  normalised_exprs=normalised_exprs[g,combat_phenotype$sample_id])
ggplot(dat, aes(x=cycle_time, y=normalised_exprs)) +
  geom_point() +
  geom_hline(yintercept=gene_means[g], linetype="dashed") +
  scale_y_continuous(limits=c(5.0, 9.1)) +
  labs(title="ENSG00000187231 (cycle normalised)")
```

```{r}
g <- "ENSG00000100485"
plot_spline_model(gene_models[[g]],
                  y=combat_nc[g,], x=cycle_time,
                  gene_name = g, plot=FALSE) +
  scale_y_continuous(limits=c(5, 7)) 
dat <- data.frame(cycle_time=cycle_time,
                  normalised_exprs=normalised_exprs[g,combat_phenotype$sample_id])
ggplot(dat, aes(x=cycle_time, y=normalised_exprs)) +
  geom_point() +
  geom_hline(yintercept=gene_means[g], linetype="dashed") +
  scale_y_continuous(limits=c(5, 7)) +
  labs(title="ENSG00000100485 (cycle normalised)")
```


-----

# Save data

```{r}
unique_id <- "2021-03-23_k30_g4"
saveRDS(gene_models, file=sprintf("data/gene_models/gene_models_%s.rds", unique_id))
saveRDS(normalised_exprs, file=sprintf("data/tidy_data/cycle_normalised_exprs_%s.rds",
                                       unique_id))
```

-----

# Session Info

```{r}
devtools::session_info()
```
