---
title: "Batch Effects EDA"
author: "Jessica Chung"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: true
    toc_depth: 4
---

Exploratory analysis of batch effects in the microarray data.

Try combat with and without cycle stage, and compare.


```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=TRUE, message=TRUE, error=TRUE, echo=TRUE, results="hold")
knitr::opts_knit$set(root.dir = "..")
options(digits=4)
Sys.setenv(TZ = "Australia/Melbourne")
```

-----

# Load

```{r message=FALSE}
library(tidyverse)
library(ggplot2)
library(sva)
library(limma)
library(edgeR)
```

### Load data

```{r}
# Microarray data
exprs <- readRDS("data/array_data/combined_bg_rsn_exprs.rds")
pval <- readRDS("data/array_data/combined_pval.rds")

# Phenotype data
phenotype <- readRDS("data/tidy_data/sample_info.rds")

# Batch data
batch_data <- readRDS("data/array_data/batch_data.rds")
```

```{r}
stopifnot(colnames(exprs) == colnames(pval))
```

-----

# Wrangle

```{r}
phenotype <- phenotype %>%
  mutate(study=ifelse(str_detect(sample_id, "^URS"), "study_2", "study_1"))
urs_array_samples <- phenotype %>% 
  filter(! is.na(array_data)) %>% 
  filter(study == "study_2") %>%
  pull(sample_id)
```

```{r}
# Change X# names of URS samples to URS###
names_to_change <- batch_data %>% filter(study == "study_2") %>% .$sample_id
tmp <- str_match(names_to_change, "^X([0-9]+)(_?\\w)?$")
tmp[,2] <- sprintf("%03d", as.numeric(tmp[,2]))
tmp[,3][is.na(tmp[,3])] <- ""
new_names <- paste0("URS", tmp[,2], tmp[,3])
new_name_lookup <- new_names
names(new_name_lookup) <- names_to_change
stopifnot(urs_array_samples %in% new_name_lookup)
new_name_lookup <- new_name_lookup[new_name_lookup %in% urs_array_samples]
```

```{r}
urs_indices <- colnames(exprs) %in% names(new_name_lookup)
colnames(exprs)[urs_indices] <- new_name_lookup[colnames(exprs)[urs_indices]]
colnames(pval)[urs_indices] <- new_name_lookup[colnames(pval)[urs_indices]]

urs_indices <- batch_data$sample_id %in% names(new_name_lookup)
batch_data$sample_id[urs_indices] <-
  new_name_lookup[batch_data$sample_id[urs_indices]]
```

Combine phenotype cycle stages and batch data cycle stages. Cycle stage from
batch data may not be reliable.

```{r}
cycle_stage_df <- merge(phenotype, batch_data, by="sample_id", 
                        suffixes=c("", ".old"), all.y = TRUE) %>%
  select(sample_id, contains("cycle_stage")) %>%
  mutate(cycle_stage=ifelse(is.na(cycle_stage), cycle_stage.old, cycle_stage)) %>%
  filter(!is.na(cycle_stage)) %>%
  select(-cycle_stage.old)
```


-----

# Check batch effects

Note there are no samples in batch 4 due to no RedCap data.

```{r}
# Use all array data from study 1
study_1_batch_data <- batch_data %>% filter(study == "study_1")
study_1_batch_data$batch %>% table
```

```{r}
# Remove replicates and control samples, and add cycle stage info
s1_pheno <- study_1_batch_data %>% 
  filter(str_detect(sample_id, "^X\\d+$")) %>% 
  select(sample_id, batch, sentrix_barcode) %>%
  merge(cycle_stage_df, by="sample_id", all.x=TRUE)
```


```{r}
# Check cycle stage distribution for each batch
s1_pheno %>% filter(cycle_stage %in% 1:7) %>% 
  with(table(batch, cycle_stage))
```


## PCA

```{r}
# Look at all batches except 4, coloured by cycle stage
subset <- s1_pheno %>% filter(batch != 4) %>% 
  filter(cycle_stage %in% 1:7)
pca <- prcomp(t(exprs[,subset$sample_id]))$x[,1:3]
dat <- cbind(subset, pca)
ggplot(dat, aes(x=PC1, y=PC2, color=cycle_stage)) +
  geom_point()
```

```{r}
# PCA looking at a particular cycle stage
# If using all cycle stages, array batch effect is drowned out
subset <- s1_pheno %>% filter(cycle_stage == 6)
pca <- prcomp(t(exprs[,subset$sample_id]))$x[,1:3]
dat <- cbind(subset, pca)
ggplot(dat, aes(x=PC1, y=PC2, color=batch)) +
  geom_point()
```

## Pairwise batch comparison

Pairwise comparison of batches.

```{r}
# Randomly use 10 samples from each batch
set.seed(20180116)
n_sample <- 10
batches <- s1_pheno$batch %>% unique %>% sort
dge_samples <- c()
for (b in batches) {
  subset <- s1_pheno %>% filter(batch == b)
  dge_samples <- c(dge_samples, sample(subset$sample_id, n_sample))
}

# Get array data
dge_pheno <- s1_pheno[match(dge_samples, s1_pheno$sample_id),]
dge_probes <- rowMeans(pval[,dge_pheno$sample_id] < 0.05) %>% 
  magrittr::is_greater_than(0.9) %>%
  names(.)[.]
dge_y <- exprs[dge_probes,dge_pheno$sample_id]
dge_results <- matrix(0, nrow=length(batches), ncol=length(batches))

# Pairwise comparisons
for (i in seq_along(batches)) {
  b1 <- batches[i]
  for (j in seq_along(batches)) {
    b2 <- batches[j]
    if (b1 == b2) {
      next
    }
    pheno <- dge_pheno %>% filter(batch %in% c(b1, b2))
    y <- dge_y[,pheno$sample_id]
    batch <- pheno$batch %>% factor
    design <- model.matrix(~batch)
    fit <- lmFit(y, design) %>% eBayes
    top <- topTable(fit, coef=2, n=Inf)
    dge_results[i, j] <- sum(top$adj.P.Val < 0.01)
  }
}

# View results
dge_results %>% magrittr::set_colnames(batches) %>% 
  magrittr::set_rownames(batches)
```


## Sentrix barcode batch effect

See if barcodes have an effect within each batch.

```{r}
# How confounded are barcodes with cycle stage?
study_1_batch_data %>% filter(batch == 7) %>%
  with(table(sentrix_barcode, cycle_stage))
```

For each batch, for each sentrix barcode, only use samples in the
proliferative stage.

```{r}
# Get all barcodes with at least three samples. Only use samples in prolif.
n_sample <- 3
subset_pheno <- s1_pheno %>% filter(cycle_stage %in% 2:4)
batches <- subset_pheno$batch %>% unique %>% sort
dge_results <- list()

# Get array data
dge_probes <- rowMeans(pval[,dge_pheno$sample_id] < 0.05) %>% 
  magrittr::is_greater_than(0.9) %>%
  names(.)[.]
dge_y <- exprs[dge_probes,subset_pheno$sample_id]

# Pairwise comparisons
for (b in batches) {
  barcodes <- subset_pheno %>% filter(batch == b) %>% .$sentrix_barcode %>%
    table %>% magrittr::is_weakly_greater_than(n_sample) %>% names(.)[.]
  dge_results[[b]] <- matrix(0, nrow=length(barcodes), ncol=length(barcodes))
  
  for (i in seq_along(barcodes)) {
    b1 <- barcodes[i]
    for (j in seq_along(barcodes)) {
      b2 <- barcodes[j]
      if (b1 == b2) {
        next
      }
      pheno <- subset_pheno %>% filter(sentrix_barcode %in% c(b1, b2))
      y <- dge_y[,pheno$sample_id]
      batch <- pheno$sentrix_barcode %>% factor
      design <- model.matrix(~batch)
      fit <- lmFit(y, design) %>% eBayes
      top <- topTable(fit, coef=2, n=Inf)
      dge_results[[b]][i, j] <- sum(top$adj.P.Val < 0.05)
    }
  }
  rownames(dge_results[[b]]) <- barcodes
  colnames(dge_results[[b]]) <- barcodes
  
}
```

```{r}
# Only print batches with non-zero values
for (b in batches) {
  if (! all(dge_results[[b]] == 0)) {
    print(b)
    print(dge_results[[b]])
  }
}
```

Batch 4 and batch 7 has some differences between sentrix barcodes.

```{r}
# Look at the comparison with the most differences in detail
cat("200536850048 vs 200536980011")
pheno <- subset_pheno %>% filter(sentrix_barcode %in% c("200536850048", "200536980011"))
y <- dge_y[,pheno$sample_id]
batch <- pheno$sentrix_barcode %>% factor
design <- model.matrix(~batch)
fit <- lmFit(y, design) %>% eBayes
top <- topTable(fit, coef=2, n=Inf)
hist(top$P.Value, breaks=100)
```



```{r}
# Look at the comparison with the most differences in detail
cat("6898481038 vs 6898481037")
pheno <- subset_pheno %>% filter(sentrix_barcode %in% c("6898481038", "6898481037"))
y <- dge_y[,pheno$sample_id]
batch <- pheno$sentrix_barcode %>% factor
design <- model.matrix(~batch)
fit <- lmFit(y, design) %>% eBayes
top <- topTable(fit, coef=2, n=Inf)
hist(top$P.Value, breaks=100)
```


## PCA within batches

Correct for cycle stage using residuals.

```{r}
# All samples
# Fit linear model and get residuals
subset <- s1_pheno %>% filter(batch %in% 1:7) %>% filter(cycle_stage %in% 1:7)
design <- model.matrix(~cycle_stage, subset)
fit <- lmFit(exprs[,subset$sample_id], design=design)
Xbeta <- design %*% t(fit$coefficients)
rownames(Xbeta) <- subset$sample_id
residuals <- exprs[,subset$sample_id] - t(Xbeta)

# PCA plot with all samples
pca <- prcomp(t(residuals))$x[,1:4]
dat <- cbind(subset, pca)
ggplot(dat, aes(x=PC1, y=PC2, color=batch)) +
  geom_point()
```


```{r}
# Batch 4
# Fit linear model and get residuals
subset <- s1_pheno %>% filter(batch %in% 4) %>% filter(cycle_stage %in% 1:7)
design <- model.matrix(~cycle_stage, subset)
fit <- lmFit(exprs[,subset$sample_id], design=design)
Xbeta <- design %*% t(fit$coefficients)
rownames(Xbeta) <- subset$sample_id
residuals <- exprs[,subset$sample_id] - t(Xbeta)

# PCA plot with all samples
pca <- prcomp(t(residuals))$x[,1:4]
dat <- cbind(subset, pca)
ggplot(dat, aes(x=PC1, y=PC2, color=sentrix_barcode)) +
  geom_point()
```

```{r}
# Batch 7
# Fit linear model and get residuals
subset <- s1_pheno %>% filter(batch %in% 7) %>% filter(cycle_stage %in% 1:7)
design <- model.matrix(~cycle_stage, subset)
fit <- lmFit(exprs[,subset$sample_id], design=design)
Xbeta <- design %*% t(fit$coefficients)
rownames(Xbeta) <- subset$sample_id
residuals <- exprs[,subset$sample_id] - t(Xbeta)

# PCA plot with all samples
pca <- prcomp(t(residuals))$x[,1:4]
dat <- cbind(subset, pca)
ggplot(dat, aes(x=PC1, y=PC2, color=sentrix_barcode)) +
  geom_point()
```


```{r}
# Two barcodes
# Fit linear model and get residuals
subset <- s1_pheno %>% 
  filter(sentrix_barcode %in% c("200536850048", "200536980011")) %>% 
  filter(cycle_stage %in% 1:7)
design <- model.matrix(~cycle_stage, subset)
fit <- lmFit(exprs[,subset$sample_id], design=design)
Xbeta <- design %*% t(fit$coefficients)
rownames(Xbeta) <- subset$sample_id
residuals <- exprs[,subset$sample_id] - t(Xbeta)

# PCA plot with all samples
pca <- prcomp(t(residuals))$x[,1:4]
dat <- cbind(subset, pca)
ggplot(dat, aes(x=PC1, y=PC2, color=sentrix_barcode)) +
  geom_point()
```

Only a few cases where sentrix barcodes have an observable effect.

Only use batches for batch correction, and not sentrix barcodes.

-----

# Combat batches 1-7

Remove batch effect with cycle stage as a covariate.

```{r}
# Get expression matrix
y <- exprs[,s1_pheno$sample_id]

# Get batch info
batch <- s1_pheno$batch %>% factor
table(batch)

# Combat covariates
mod <- model.matrix(~cycle_stage, data=s1_pheno)
head(mod)

# Combat normalisation
combat_with_cycle_exprs <- ComBat(y, batch=batch, mod=mod)
```

Remove batch effect without cycle stage as a covariate.

```{r}
# Get expression matrix
y <- exprs[,s1_pheno$sample_id]

# Get batch info
batch <- s1_pheno$batch %>% factor
table(batch)

# Combat covariates
mod <- model.matrix(~1, data=s1_pheno)

# Combat normalisation
combat_without_cycle_exprs <- ComBat(y, batch=batch, mod=mod)
```

-----

# Post-batch correction analysis

## PCA

```{r}
# Look at all batches
subset <- s1_pheno %>% filter(cycle_stage %in% 1:7)
pca <- prcomp(t(combat_with_cycle_exprs[,subset$sample_id]))$x[,1:3]
dat <- cbind(subset, pca)
ggplot(dat, aes(x=PC1, y=PC2, color=cycle_stage)) +
  geom_point()
```

```{r}
# Look at all batches
subset <- s1_pheno %>% filter(cycle_stage %in% 1:7)
pca <- prcomp(t(combat_without_cycle_exprs[,subset$sample_id]))$x[,1:3]
dat <- cbind(subset, pca)
ggplot(dat, aes(x=PC1, y=PC2, color=cycle_stage)) +
  geom_point()
```


```{r}
# PCA looking at a particular cycle stage
subset <- s1_pheno %>% filter(cycle_stage == 6)
pca <- prcomp(t(combat_without_cycle_exprs[,subset$sample_id]))$x[,1:3]
dat <- cbind(subset, pca)
ggplot(dat, aes(x=PC1, y=PC2, color=batch)) +
  geom_point()
```

## Pairwise batch comparisons

Redo pairwise batch comparisons using combat without cycle stage

```{r}
# Randomly use 10 samples from each batch
set.seed(20180116)
n_sample <- 10
batches <- s1_pheno$batch %>% unique %>% sort
dge_samples <- c()
for (b in batches) {
  subset <- s1_pheno %>% filter(batch == b)
  dge_samples <- c(dge_samples, sample(subset$sample_id, n_sample))
}

# Get array data
dge_pheno <- s1_pheno[match(dge_samples, s1_pheno$sample_id),]
dge_probes <- rowMeans(pval[,dge_pheno$sample_id] < 0.05) %>% 
  magrittr::is_greater_than(0.9) %>%
  names(.)[.]
dge_y <- combat_without_cycle_exprs[dge_probes,dge_pheno$sample_id]
dge_results <- matrix(0, nrow=length(batches), ncol=length(batches))

# Pairwise comparisons
for (i in seq_along(batches)) {
  b1 <- batches[i]
  for (j in seq_along(batches)) {
    b2 <- batches[j]
    if (b1 == b2) {
      next
    }
    pheno <- dge_pheno %>% filter(batch %in% c(b1, b2))
    y <- dge_y[,pheno$sample_id]
    batch <- pheno$batch %>% factor
    design <- model.matrix(~batch)
    fit <- lmFit(y, design) %>% eBayes
    top <- topTable(fit, coef=2, n=Inf)
    dge_results[i, j] <- sum(top$adj.P.Val < 0.01)
  }
}

# View results
dge_results %>% magrittr::set_colnames(batches) %>% 
  magrittr::set_rownames(batches)
```

## Cycle stage differential expression

Compare results between combat with and without cycle stage.

Expect combat without cycle stage will reduce the effects from cycle stage,
but by how much?

```{r}
# Randomly use 12 samples from each batch
# Exclude cycle stage 2 due to too few samples
set.seed(20180116)
n_sample <- 12
cycles <- c(1, 3:7)
dge_samples <- c()
for (b in cycles) {
  subset <- s1_pheno %>% filter(cycle_stage == b)
  dge_samples <- c(dge_samples, sample(subset$sample_id, n_sample))
}

# Get array data
dge_pheno <- s1_pheno[match(dge_samples, s1_pheno$sample_id),]
dge_probes <- rowMeans(pval[,dge_pheno$sample_id] < 0.05) %>% 
  magrittr::is_greater_than(0.9) %>%
  names(.)[.]
dge_y <- combat_without_cycle_exprs[dge_probes,dge_pheno$sample_id]
dge_results <- matrix(0, nrow=length(cycles), ncol=length(cycles))

# Pairwise comparisons
for (i in seq_along(cycles)) {
  c1 <- cycles[i]
  for (j in seq_along(cycles)) {
    c2 <- cycles[j]
    if (c1 == c2) {
      next
    }
    pheno <- dge_pheno %>% filter(cycle_stage %in% c(c1, c2))
    y <- dge_y[,pheno$sample_id]
    cycle <- pheno$cycle_stage %>% factor
    design <- model.matrix(~cycle)
    fit <- lmFit(y, design) %>% eBayes
    top <- topTable(fit, coef=2, n=Inf)
    dge_results[i, j] <- sum(top$adj.P.Val < 0.01)
  }
}

# View results
dge_results %>% magrittr::set_colnames(cycles) %>% 
  magrittr::set_rownames(cycles)
```

Redo with combat with cycle stage

```{r}
# Randomly use 12 samples from each batch
# Exclude cycle stage 2 due to too few samples
set.seed(20180116)
n_sample <- 12
cycles <- c(1, 3:7)
dge_samples <- c()
for (b in cycles) {
  subset <- s1_pheno %>% filter(cycle_stage == b)
  dge_samples <- c(dge_samples, sample(subset$sample_id, n_sample))
}

# Get array data
dge_pheno <- s1_pheno[match(dge_samples, s1_pheno$sample_id),]
dge_probes <- rowMeans(pval[,dge_pheno$sample_id] < 0.05) %>% 
  magrittr::is_greater_than(0.9) %>%
  names(.)[.]
dge_y <- combat_with_cycle_exprs[dge_probes,dge_pheno$sample_id]
dge_results <- matrix(0, nrow=length(cycles), ncol=length(cycles))

# Pairwise comparisons
for (i in seq_along(cycles)) {
  c1 <- cycles[i]
  for (j in seq_along(cycles)) {
    c2 <- cycles[j]
    if (c1 == c2) {
      next
    }
    pheno <- dge_pheno %>% filter(cycle_stage %in% c(c1, c2))
    y <- dge_y[,pheno$sample_id]
    cycle <- pheno$cycle_stage %>% factor
    design <- model.matrix(~cycle)
    fit <- lmFit(y, design) %>% eBayes
    top <- topTable(fit, coef=2, n=Inf)
    dge_results[i, j] <- sum(top$adj.P.Val < 0.01)
  }
}

# View results
dge_results %>% magrittr::set_colnames(cycles) %>% 
  magrittr::set_rownames(cycles)
```

When not using cycle stage as a covariate, expression effects from cycle stage 
are reduced, but there's still lots of signal.

Some samples have unknown cycle stages, so it's better to just keep all
study 1 samples instead of only using those with known cycle stages.

Therefore, use combat without cycle stage as a covariate.

-----

# Study 2

## Check batch effects

```{r}
study_2_batch_data <- batch_data %>%
  filter(study == "study_2")
```


```{r}
study_2_batch_data$batch %>% table
```

```{r}
# Remove replicates and control samples, and merge with sample info
s2_pheno <- merge(study_2_batch_data, phenotype, by="sample_id")

stopifnot(s2_pheno$sample_id %in% colnames(exprs))
stopifnot(! is.na(s2_pheno$batch))
```


```{r}
# Check day distribution for each batch
with(s2_pheno, table(batch, consensus_pod_urs_only))
```


## PCA

```{r}
subset <- s2_pheno %>% filter(! is.na(consensus_pod_urs_only))
pca <- prcomp(t(exprs[,subset$sample_id]))$x[,1:3]
dat <- cbind(subset, pca)
ggplot(dat, aes(x=PC1, y=PC2, color=batch, label=sample_id)) +
  geom_point() +
  geom_text(size=2, nudge_x=3)
```

Batch 6 is comprised of two separate batches.

```{r}
ggplot(dat %>% filter(batch == "6"), 
       aes(x=PC1, y=PC2, color=sentrix_barcode)) +
  geom_point()
```

```{r}
# Plot with cycle day
ggplot(dat, aes(x=PC1, y=PC2, color=consensus_pod_urs_only)) +
  geom_point()
```

## Separate batch 6

```{r}
batch_6.1_barcodes <- c("3999650006", "3999650008", "3999650067", "3999650060")
batch_6.2_barcodes <- c("3999650037", "3999650055", "3999650048", "3999650059")

s2_pheno <- s2_pheno %>% 
  mutate(batch=ifelse(sentrix_barcode %in% batch_6.1_barcodes, "6.1", batch),
         batch=ifelse(sentrix_barcode %in% batch_6.2_barcodes, "6.2", batch))
```

```{r}
subset <- s2_pheno %>% filter(! is.na(consensus_pod_urs_only))
pca <- prcomp(t(exprs[,subset$sample_id]))$x[,1:3]
dat <- cbind(subset, pca)
ggplot(dat, aes(x=PC1, y=PC2, color=batch)) +
  geom_point()
```

```{r}
with(s2_pheno, table(batch, consensus_pod_urs_only))
```


-----

# Session Info

```{r}
devtools::session_info()
```

